---
title: Creating maps with R and OpenStreetMap (2)
author: Florian Zenoni
date: '2017-11-07'
categories:
  - OpenStreetMap
  - R
tags:
  - OpenStreetMap
  - sf
  - tmap
  - osmdata
slug: creating-maps-coastlines
---

```{r, echo=FALSE}
# CONFIG
user_name <- "fzenoni" # your Git username (only needed if
# you want to deploy to GH pages)
project_name <- "coast-map" # adapt!
package_date <- "2017-11-07" # date of the CRAN snapshot that
# the checkpoint package uses
```

### GitHub

The code for the herein described process can also be freely downloaded from [https://github.com/`r user_name`/`r project_name`](https://github.com/`r user_name`/`r project_name`). 

```{r include=FALSE}
detach_all_packages <- function() {
  basic_packages_blank <-  c("stats",
    "graphics",
    "grDevices",
    "utils",
    "datasets",
    "methods",
    "base")
  basic_packages <- paste("package:", basic_packages_blank, sep = "")
  
  package_list <- search()[
    ifelse(unlist(gregexpr("package:", search())) == 1, TRUE, FALSE)]
  
  package_list <- setdiff(package_list, basic_packages)
  
  if (length(package_list) > 0)  for (package in package_list) {
    detach(package, character.only = TRUE, unload = TRUE)
    print(paste("package ", package, " detached", sep = ""))
  }
}

detach_all_packages()

# this allows multiple persons to use the same RMarkdown
# without adjusting the working directory by themselves all the time
source("scripts/csf.R")
path_to_wd <- csf() # if this - for some reason - does not work, 
# replace with a hardcoded path, like so: "~/projects/rddj-template/analysis/"
if ( is.null(path_to_wd) | !dir.exists(path_to_wd)) {
  print("WARNING: No working directory specified for current user")
} else {
  setwd(path_to_wd)
}
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# from https://mran.revolutionanalytics.com/web/packages/checkpoint/vignettes/using-checkpoint-with-knitr.html
# if you don't need a package, remove it from here (commenting is probably not sufficient)
# tidyverse: see https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/
cat("
library(backports)
library(dplyr)
library(osmdata)
library(sf)
library(knitr)
library(tmap)",
  file = "manifest.R")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# if checkpoint is not yet installed, install it (for people using this
# system for the first time)
if (!require(checkpoint)) {
  if (!require(devtools)) {
    install.packages("devtools", repos = "http://cran.us.r-project.org")
    require(devtools)
  }
  devtools::install_github("checkpoint",
    username = "RevolutionAnalytics",
    ref = "v0.3.2", # could be adapted later,
    # as of now (beginning of July 2017
    # this is the current release on CRAN)
    repos = "http://cran.us.r-project.org")
  require(checkpoint)
}
# nolint start
if (!dir.exists("~/.checkpoint")) {
  dir.create("~/.checkpoint")
}
# nolint end
# install packages for the specified CRAN snapshot date
checkpoint(snapshotDate = package_date,
  project = path_to_wd,
  verbose = T,
  scanForPackages = T,
  use.knitr = F)
rm(package_date)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
source("manifest.R")
unlink("manifest.R")
# sessionInfo()
```

## Where were we

In the previous post (click [here](https://altran-data-analytics.netlify.com/2017/10/10/2017-10-10-creating-maps-with-r-and-openstreetmap/) if you missed it), we managed to put together a relatively short chunk of code that generated the following rather nice map of Brussels Region, which includes all of its 19 municipalities.

```{r message=FALSE, warning=FALSE}
boundaries <- opq(bbox = 'Brussels, Belgium') %>%
  add_osm_feature(key = 'admin_level', value = '8') %>% 
  osmdata_sf %>% unique_osmdata
municipalities <- boundaries$osm_multipolygons
regions <- opq(bbox = 'Brussels, Belgium') %>%
  add_osm_feature(key = 'admin_level', value = '4') %>% 
  osmdata_sf %>% unique_osmdata
bxl_region <- regions$osm_multipolygons %>% filter(osm_id == '54094')

municipalities <- st_transform(municipalities, 3812)
bxl_region <- st_transform(bxl_region, 3812)
# Fix the polygons before the intersection, or it will fail
if(!all(st_is_valid(municipalities))) {
  municipalities <- municipalities %>% st_make_valid
}

neg_buffer <- st_buffer(bxl_region, -100) # in meters
bxl_municipalities_poly <- municipalities[neg_buffer, ]

# Drop the municipalities factors excluded from the latest subset operations
# or they will appear in the legend.
bxl_municipalities_poly <- bxl_municipalities_poly %>% mutate(name = droplevels(name))

tm_style_col_blind() +
  tm_shape(bxl_municipalities_poly) +
  tm_polygons(title = 'Brussels Capital Municipalities', border.col = 'grey40', col = 'name', alpha = 0.6) +
  tm_shape(bxl_region) + tm_borders(col = 'grey20', lwd = 2, alpha = 0.8) +
  tm_layout(legend.outside = TRUE, frame.double.line = TRUE) +
  tm_grid(projection = 'longlat', n.x = 5) + tm_scale_bar() +
  tm_compass(position = c('right', 'top'))
```

In that post, we discussed how a certain amount of challenges were overcome. We picked Brussels, a city that lays in mainland. As a consequence, we did not have to worry about what stands _around_ Brussels, because it wouldn't have added much information to our map. But what if instead we decided to display a city on the coast? For this purpose, let's choose a warmer place, such as Bilbao, Spain! Besides the presence of the sea, this choice will give us the chance check how to visualize Bilbao's estuary, and also possibly the flowing river.

At this stage I should specify that `tmap` gives you the chance to superimpose your queries to the OpenStreetMap map tiles, by creating an interactive HTML widget. This could be the shortest possible answer to display the region of interest in its geographical context. To showcase this let's begin once again by the fundamentals. We are going to query the administrative level 7, corresponding the the _comarca_ of Greater Bilbao (some sort of Spanish county equivalent).

```{r}
boundaries <- opq(bbox = 'Bilbao, Spain') %>%
  add_osm_feature(key = 'admin_level', value = '7') %>% 
  osmdata_sf %>% unique_osmdata
```

Now let's activate the "view" option (the default being "plot"), and let's display the single region we queried.

```{r message=FALSE}
tmap_mode('view')
qtm(boundaries$osm_multipolygons)
# Reset to default for next steps
tmap_mode('plot')
```

It's a cool solution when building websites or interactive presentations. But it's not optimal when a map needs to be printed and the programmer wants to highlight specific aspects of the map.

We know how to query boundaries with `osmdata`; nothing stops us from querying a coastline.
But first, I will define a slightly larger bounding box, so that we are mathematically sure that our boundaries and their surroundings will be correctly displayed.

```{r}
bb <- st_bbox(boundaries$osm_multipolygons)
# Defining a buffer
buffer <- 0.1
p_big <- rbind(c(bb[1] - buffer, bb[2] - buffer),
  c(bb[1] - buffer, bb[4] + buffer),
  c(bb[3] + buffer, bb[4] + buffer),
  c(bb[3] + buffer, bb[2] - buffer),
  c(bb[1] - buffer, bb[2] - buffer))
# Putting the coordinates into a squared polygon object
pol <- st_polygon(list(p_big)) %>% st_geometry
# Providing the SRID (here, unprojected lon/lat)
st_crs(pol) <- 4326
```

The correct `key` and `value` to obtain the coastline are `natural` and `coastline` (check the [wiki page](http://wiki.openstreetmap.org/wiki/Tag:natural%3Dcoastline) for more details).

```{r}
coast <- opq(bbox = st_bbox(pol)) %>%
  add_osm_feature(key = 'natural', value = 'coastline') %>%
  osmdata_sf %>% unique_osmdata

qtm(coast$osm_lines)
```

Interesting already! We got Bilbao's coastline, some tiny islands here and there, as well as the estuary.
We can quickly superimpose what we have.

```{r}
qtm(boundaries$osm_multipolygons) + qtm(coast$osm_lines)
```

Is there more to it? Can we see the rest of the river? To find out, I am going to query any OSM object carrying the key "waterway".

```{r}
water <- opq(bbox = st_bbox(pol)) %>%
  add_osm_feature(key = 'waterway') %>%
  osmdata_sf %>% unique_osmdata
```

After a quick inspection, it seems that the watercourses are represented as polygons. Let's add them to the rest.

```{r}
qtm(boundaries$osm_multipolygons) + qtm(coast$osm_lines) + qtm(water$osm_polygons)
```

It is a very rough visualization, but as one can guess, almost all the elements are here.

Now, no map of a coastal city would be complete without appropriate colors for sea and land. If land and sea were univoquely represented by polygons, this would be an easy task. Instead, all we have, at least for the coastline, is a `data.frame` of `LINESTRING` objects. The strategy would then be to cut the polygon frame in two parts, according to the coast acting as a "blade". Luckily for us, the `sf` package includes many of the functions available in PostGIS technology. This time, it's the time for `st_split` to shine.

_N.B.: Things did not exactly go as expected for this blog post. I still pay the price for some lack of experience. In the following lines, if you know a better solution, or understand what is going wrong, don't hesitate to write in the comments!_

First, we define the blade by taking the coastline, making a union of all the `data.frame` rows, thus obtaining a `MULTILINESTRING`. Since after inspection, that object is truly a single string, we use the appropriate function to merge everything. This is very important, because `st_split` does not support cutting a polygon with a multilinestring (as reported in the PostGIS [doc](https://postgis.net/docs/ST_Split.html)).

```{r}
blade <- coast$osm_lines %>% st_union %>% st_line_merge
```

We use this `blade` to cut the squared polygon we defined earlier.

```{r}
multipol <- st_split(st_geometry(pol), st_geometry(blade))
```

The result is a `GEOMETRYCOLLECTION`. In principle, one could access the different elements of the collection (in this case, two) by using double square brackets. For a reason I could not understand, in this case it works only for the first element, which I then cast appropriately.

```{r}
land <- st_cast(multipol[[1]], 'POLYGON') %>% st_geometry %>% st_sf
st_crs(land) <- 4326
```

This is why, instead of calling the second polygon with `multipol[[2]]`, I am forced to make a different kind of operation. Looking at the bright side, this is yet another display of possible `sf` geometrical operations.

```{r}
sea <- st_difference(pol, land) %>% st_geometry %>% st_sf
st_crs(sea) <- 4326
```

We are now ready to display everything, similarly to what was done in the last post. Note the `is.master = T` activated option, which calibrates the borders of the final picture according to Bilbao's boundaries, instead of the polygon called first (in this case, `land`).

```{r}
tm_shape(land) + tm_polygons(col = 'bisque3', alpha = 0.8) +
  tm_shape(boundaries$osm_multipolygons, is.master = T) + tm_polygons() +
  tm_shape(sea) + tm_polygons(col = 'skyblue', alpha = 0.8) +
  tm_shape(water$osm_polygons) + tm_polygons(col = 'skyblue', alpha = 0.8) +
  tm_grid(n.x = 5, projection = "longlat") +
  tm_layout(legend.outside = TRUE, title = 'Bilbao, Spain', frame.double.line = TRUE) +
  tm_scale_bar() + tm_compass(position = c('right', 'top'))
```
